---
title: "16S-workshop"
author: "Hardik I Parikh"
date: "4/18/2018"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
      number_sections: true
---
```{r setup, include=FALSE}
#### Setup environment

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
library(phyloseq)
library(vegan)
library(ape)
library(ggpubr)
library(RColorBrewer)
theme_set(theme_bw())
set.seed(12345)
```


This markdown outlines instructions for visualization and analysis of OTU-clustered amplicon sequencing data, primarily using the _phyloseq_ package.  

### Prerequisites

- R basics
- Data manipulation with dplyr and `%>%`
- Data visualization with ggplot2


#### R packages 

**_CRAN packages_**

- [tidyverse (readr, dplyr, ggplot2)](https://www.tidyverse.org/packages/)
- [magrittr](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html)
- [reshape2](https://cran.r-project.org/web/packages/reshape2/reshape2.pdf)
- [vegan](https://cran.r-project.org/web/packages/vegan/vignettes/intro-vegan.pdf)
- [ape](https://cran.r-project.org/web/packages/ape/ape.pdf)
- [ggpubr](http://www.sthda.com/english/rpkgs/ggpubr/)
- [RColorBrewer](https://cran.r-project.org/web/packages/RColorBrewer/RColorBrewer.pdf)

**_Bioconductor packages_**

- [phyloseq](https://joey711.github.io/phyloseq/)
- [DESeq2](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)


#### Required Data Files
We will use the output files generated during the sequence processing steps.

- OTU table (abundance table) : `relman2017_samples.otu_table.txt`
- Taxonomy table : `relman2017_samples.tax_table.txt`
- Sample metadata : `relman2017_samples.sample_data.txt`
- OTU phylogenetic tree : `relman2017_samples.rep_set.tre`


### Load data
Let's read in the data files using `read_tsv()` function of **readr** package 
```{r}
dataDir = "~/Documents/16S-workshop/output/"
# import OTU table
otu <- read_tsv(paste(dataDir, "relman2017_samples.otu_table.txt", sep="/"))
otu
# import Taxonomy table
taxonomy <- read_tsv(paste(dataDir, "relman2017_samples.tax_table.txt", sep="/"))
taxonomy
# import Sample metadata
metadata <- read_tsv(paste(dataDir, "relman2017_samples.sample_data.txt", sep="/"))
metadata
# import phylogenetic tree
phytree <- read.tree(paste(dataDir, "relman2017_samples.rep_set.tre", sep="/"))
phytree
```


### Create phyloseq object
Read More: [Importing data into phyloseq](https://joey711.github.io/phyloseq/import-data.html)
```{r}

# create OTU table
otu <- otu %>%
  as.data.frame() %>%
  column_to_rownames("OTUId")
OTU <- otu_table(otu, taxa_are_rows = TRUE)  

# create TAX table
taxonomy <- taxonomy %>%
  as.data.frame() %>%
  column_to_rownames("X1") %>%
  as.matrix()
TAX <- tax_table(taxonomy)

# create sample metadata
metadata <- metadata %>%
  as.data.frame() %>%
  mutate_if(sapply(metadata, is.character), as.factor) %>%
  column_to_rownames("sample") 
SDATA <- sample_data(metadata)

# create the phyloseq object
physeq <- phyloseq(OTU, TAX, SDATA, phytree)
```


### Data pruning
Read More: [(Pre)Processing Data](https://joey711.github.io/phyloseq/preprocess.html)  
_Phyloseq_ package provides a plethora of functions for filtering, subsetting and merging abundance data. It is beyond the scope of this workshop to discuss their usage in detail and downstream implications of chosen threshold. I strongly recommend referring the _phyloseq_ vignette for answers!  
For this tutorial, we will remove taxa that have read count less than 100 in at least 10% of samples. This protects againsts OTUs with small mean and trivially large C.V.  
*Do not use these measures to filter your dataset!*
```{r}
# For this tutorial, we will only look into high-abundance/high-prevelance OTUs
physeq.f <- filter_taxa(physeq, 
                        function(x) sum(x >= 100) > (0.01*length(x)), 
                        prune = TRUE)
```


### Stacked Bars
Read More: [Phyloseq bar plots](https://joey711.github.io/phyloseq/plot_bar-examples.html)  

Let's plot phylum-level abundances using the `plot_bar()` function of _phyloseq_
```{r}

plot_bar(physeq.f, "phylum", fill="phylum", facet_grid = term_vs_preterm_delivery~race) + 
  geom_bar(aes(color=phylum, fill=phylum), stat="identity", position = "stack") 

# Go over the phyloseq tutorials to explore additional features
```


Next, we will plot stacked bars by clustering samples on their bray-curtis dissimilarities. 

```{r}

# transform counts to relative abundance
physeq.f.ra <- transform_sample_counts(physeq.f, function(x) x*100/sum(x))

# agglomerate counts at genus-level
physeq.f.ra.genus <- tax_glom(physeq.f.ra, "genus")

# next get the otu_table
propData <- as.data.frame(t(otu_table(physeq.f.ra.genus)))

# get melted dataframe
plotDF <- propData %>%
  rownames_to_column(var="sample") %>%
  melt() %>%
  magrittr::set_names(c("sample", "taxa", "relab"))
  
# add taxonomy to plotting DF
taxonomy.2 <- taxonomy %>%
  as.data.frame() %>%
  rownames_to_column(var="taxa")
plotDF <- left_join(plotDF, taxonomy.2, by="taxa") %>%
  select(sample, taxa, relab, genus)

# aggregate all unclassified taxa levels
tot_unc <- plotDF %>%
  group_by(sample) %>%
  filter(grepl("unclassified", genus)) %>%
  summarise(relab=sum(relab)) %>%
  mutate(taxa = "unclassified", genus = "unclassified") 

# add the unclassified aggregated rel ab to plotting data frame
plotDF <- plotDF %>%
  filter(!grepl("unclassified", genus)) %>%
  rbind(tot_unc)

# add metadata to plotting DF
metadata.2 <- metadata %>%
  rownames_to_column(var="sample")
plotDF <- left_join(plotDF, metadata.2, by="sample")

# lets plot stacked bars using ggplot2
mycolors <- rev(colorRampPalette(brewer.pal(10, "Paired"))(length(unique(plotDF$genus))))
ggplot(plotDF, aes(sample, relab, fill=genus)) + 
  geom_bar(stat="identity", position = "stack") + 
  scale_fill_manual(values = mycolors) + 
  theme(legend.position = "bottom")

# sort samples on bray-curtis distances
bcdist <- vegdist(propData, method="bray")
hclustBC <- hclust(bcdist, method="ward.D2")
# set sample factor levels 
plotDF$sample <- factor(plotDF$sample, levels = hclustBC$labels[c(hclustBC$order)])

# plot again
ggplot(plotDF, aes(sample, relab, fill=genus, group=relab)) + 
  geom_bar(stat="identity", position = "stack") + 
  scale_fill_manual(values = mycolors) + 
  theme(legend.position = "bottom")

```


### Alpha Diversity
Read More: [Phyloseq Alpha diversity](https://joey711.github.io/phyloseq/plot_richness-examples.html)  

Alpha diversity refers to community _richness_, i.e. how many different _types_ of organisms are present, and _evenness_, i.e. how even/uneven the distribution of species abundance is, in a sample. Alpha-diversity analyses are useful for examining patterns of dominance, rarity and community complexity.

Plot alpha diversity using the `plot_richness()` function
```{r, fig.width=9, fig.height=4.5}
# alpha diversity is measured on count data in phyloseq
plot_richness(physeq.f,
              x = "race",
              color = "term_vs_preterm_delivery",
              measures = c("Observed", "Shannon", "InvSimpson")) + 
  geom_boxplot()
```

In this section, we will calculate the same measures of alpha diversity using the *vegan* package, followed by comparing means within groups using *ggpubr* package

```{r, fig.width=9, fig.height=4.5}
# calculating alpha diversity on proportional data 
# R package: vegan

propData <- as.data.frame(t(otu_table(physeq.f.ra)))

# calculate shannon index, manipulate data frame for plotting
adivDF.shannon <- as.data.frame(diversity(propData, index = "shannon")) %>%
  rownames_to_column(var="sample") %>%
  magrittr::set_names(c("sample", "value")) %>%
  mutate(index = "shannon")

# calculate invsim index, manipulate data frame for plotting
adivDF.invsim <- as.data.frame(diversity(propData, index = "invsim")) %>%
  rownames_to_column(var="sample") %>%
  magrittr::set_names(c("sample", "value")) %>%
  mutate(index = "invsim")

# create plotting dataframe
adivDF <- rbind(adivDF.shannon, adivDF.invsim) %>%
  left_join(metadata.2, by="sample")

# plot using ggpubr
ggboxplot(adivDF, "term_vs_preterm_delivery", "value", 
          color = "term_vs_preterm_delivery", palette = "jco",
          add = "jitter", outlier.shape=NA) + 
  facet_grid(~index+race) + 
  stat_compare_means(label = "p.format") 
```


### Beta Diversity, Ordinations

Read More: [Ordination Plots](https://joey711.github.io/phyloseq/plot_ordination-examples.html)  

Beta diversity refers to between sample diversity, summerize how similar/dissimilar two samples are. The taxa abundances in each sample gets compared to every other sample in the dataset, generating a distance matrix, which can be visualized using Principal Coordinate Analysis.

*Bray-Curtis distances* 
```{r}
# first calculate bray-curtis distance
physeq.f.distBC <- distance(physeq.f, method="bray")
physeq.f.distBC.ord <- ordinate(physeq.f, method = "PCoA", distance = physeq.f.distBC)

# color by metadata
plot_ordination(physeq.f, physeq.f.distBC.ord, color="race") + geom_point() 
plot_ordination(physeq.f, physeq.f.distBC.ord, color="term_vs_preterm_delivery") + geom_point() 
```

*UniFrac distances*
```{r}
# first calculate unifrac distance
physeq.f.distUF <- distance(physeq.f, method="unifrac")
physeq.f.distUF.ord <- ordinate(physeq.f, method = "PCoA", distance = physeq.f.distUF)

# color by metadata
plot_ordination(physeq.f, physeq.f.distUF.ord, color="race") + geom_point() 
plot_ordination(physeq.f, physeq.f.distUF.ord, color="term_vs_preterm_delivery") + geom_point() 
```


### DESeq2 

_**Which taxa are important?**_  
The ordination plots reveal microbiome levels shifts within groups-of-interest. A univariate analysis of individual taxa abudances can be performed using DESeq2, to measure _significant fold changes_.  

`phyloseq_to_deseq2()` function provides a convenient function to convert `phyloseq` object to `DESeq2DataSet` class.

```{r}
library(DESeq2)

# convert to DESeq2's data class
diagdds <- phyloseq_to_deseq2(physeq.f, ~ term_vs_preterm_delivery)

# perform the testing
diagdds <- DESeq(diagdds, test="Wald", fitType="parametric")

# look at the results 
res <- results(diagdds, cooksCutoff = FALSE)
sigtab <- res[which(res$padj < 0.01), ]
sigtab <- cbind(as(sigtab, "data.frame"), as(tax_table(physeq.f)[rownames(sigtab), ], "matrix"))
head(sigtab, n = 20)
```

Popular biomarker discovery tools, utilizing multivariate analysis:  
- [LEfSe](https://bitbucket.org/biobakery/biobakery/wiki/lefse)  
- [Indicator Analysis](https://cran.r-project.org/web/packages/indicspecies/vignettes/indicspeciesTutorial.pdf)


